---
title: "Statistical report: Heart Failure treatment patterns in real world"
date: "`r Sys.Date()`"
cover: static/ki_logo_vert_rgb.png
editor: source
format:
  pdf:
    documentclass: scrartcl
    template-partials:
      - "static/before-body.tex"
      - "static/_titlepage.tex"
    include-in-header: 
      - "static/in-header.tex"
    toc: true
    toc-depth: 3
    toc-title: Contents
    lof: true
    lot: true
    cap-location: top
    number-sections: true
    colorlinks: false
    keep-tex: false
bibliography: static/references.bib
csl: static/bmj.csl
link-citations: true
link-bibliography: true
nocite: '@*'
knitr:
  opts_chunk: 
    dev: cairo_pdf
    collapse: true
    comment: "" 
    echo: FALSE
    include: TRUE
    warning: FALSE
    message: FALSE
    fig.pos: "H"
    fig.path: "../output/figs/"
    fig.height: 6
    fig.width: 10
    R.options:
      knitr.graphics.auto_pdf: true
---

```{r}
#| label: set-up-load-data
options(knitr.kable.NA = "")

# load packages, globals and project specific functions
source(here::here("setup/setup.R"))

# load data
load(here("data/clean-data/rsdata.RData"))

# load workbook to write tables to Excel
wb <- loadWorkbook(here("output/tabs/tables.xlsx"))
sheets <- names(wb)

# load pptx file with figs
figs <- officer::read_pptx(path = here::here("output/figs/figs.pptx"))
output <- F
```           

\newpage

# Data

## Data sources

The Swedish Heart Failure Registry (SwedeHF) is an ongoing nationwide quality register started in 2000, 
that includes in- and out-hospital patients with HF, regardless of EF and previous history of HF[@swedehf]. 
Around 80 variables are recorded at 
discharge from hospital or at the outpatient visit, i.e. index date, and 
entered into an electronic database managed by the Uppsala Clinical Research 
Center (Uppsala, Sweden). Up to April 2017 the only inclusion criterion was a 
clinical diagnosis of HF, which was thereafter defined according to the ICD-10 
codes I50.0, I50.1, I50.9, I42.0, I42.6, I42.7, I25.5, I11.0, I13.0, I13.2. 
In 2022, 69 of 76 Swedish hospitals enrolled patients in SwedeHF, which had a 32% coverage of the prevalent HF population in Sweden[@annualreport]. 

SwedeHF was linked with national Swedish registries (SHFDB version 4.2.1) through the 
Swedish personal identity number (PIN)[@pin]: 

- The National Prescribed Drug Register (The National Board of Health and Welfare) for information on medications
- The National Patient Register (The National Board of Health and Welfare) for additional comorbidities and outcomes
- The Cause of Death Register[@dors] (The National Board of Health and Welfare) for outcomes death
- Longitudinal integrated database for health insurance and labour market studies (LISA)[@lisa] 
and the Register of the Total Population (Statistics Sweden) for socio-economic factors

## Ethics and informed consent

The study was approved by the Swedish Ethical Review Authority. 
Individual consent was not required, but patients were informed of entry into SwedeHF and able to opt‐out.

## Inclusion/exclusion criteria

```{r}
#| label: tbl-flow
#| tbl-cap: Flowchart
#| tbl-pos: "H"

flow <- flow %>%
  mutate(
    `Inclusion/exclusion criteria` = sanitize_text(`Inclusion/exclusion criteria`),
    `Inclusion/exclusion criteria` = str_replace_all(`Inclusion/exclusion criteria`, fixed("$>$="), "$\\geq$"),
    `Inclusion/exclusion criteria` = str_replace_all(`Inclusion/exclusion criteria`, fixed("$<$="), "$\\leq$")
  )

default_kable(flow, escape = F)
```

First patient included `r min(rsdata$shf_indexdtm)` and last patient included `r max(rsdata$shf_indexdtm)` 
from approximately `r rsdata %>% distinct(shf_centre) %>% count() %>% pull(n)` enrolling centers. 

## Definitions

### Baseline

Baseline is date of visits or date of discharge for the visit/hospitalization recorded in SwedeHF. 

### Follow-up

Data were censored at `r global_followup_year` years follow-up, 2023-12-31 or emigration from Sweden. 
In the survival analyses censoring was also performed at death from other causes than the event. 

\clearpage

### Variable defintions

{{< include src/vars.qmd >}}

More information on data sources and definitions are found https://kiheartfailure.github.io/shfdb4/.

### Medication

```{r}
#| label: tbl-meds
#| tbl-cap: Medications including molecules
#| tbl-pos: "H"

tab <- lmvarsub %>%
  group_by(med) %>%
  mutate(
    c = 1:n(),
    med = if_else(c == 1, med, "")
  ) %>%
  ungroup() %>%
  select(med, sub, atc)
default_kable(tab, col.names = c("Medication", "Molecule", "ATC code"), scale_down = F)
```

#### Dose

The dose taken by the patient was evaluated using the strength of the tablet prescribed x the number of pills the patients was prescribed to take each day. The latter is written in free text and an algorithm was derived to extract the dose prescribed. If there was no text or it was not possible to interpret, the number of tablets / day was assumed to be the mean number of pills taken per day for the posts with non-missing values with the same ATC code.

#### Adherence

Adherence is defined as the number of days to the next dispensation (within each medication) / the number of pills prescribed per day during the following year after baseline. Possible discontinuation will therefore be seen also in low adherence. If the percent is > 100 the adherence is assumed to be 100.  

#### Discontinuation 

Discontinuation is defined as no prescription at 6 months and 1 year $\pm$ 2 months. Patients who do not have follow-up until 6 months + 2 months and 1 year + 2 months are excluded from the respective calculations. 

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string` [@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with care.

## Missing data

Missing data was imputed with multiple imputation (n = 10) using mice [@mice] 
and Rubin’s rules was used for combining estimates and standard errors across imputed data sets. 
Variables included in the model are indicated in @tbl-base. 

\clearpage

\blandscape

# Results

```{r}
med <- rsdata %>%
  summarise(
    med = fn(median(shf_age), dig = 0),
    q1 = fn(quantile(shf_age, probs = 0.25), dig = 0),
    q3 = fn(quantile(shf_age, probs = 0.75), dig = 0)
  ) %>%
  mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
  pull(out)

percfem <- rsdata %>%
  count(shf_sex) %>%
  mutate(perc = fn(n / sum(n) * 100, 0)) %>%
  filter(shf_sex == "Female") %>%
  pull(perc)

medsurv <- rsdata %>%
  summarise(
    med = fn(median(sos_outtime_death / 30.55), dig = 0),
    mean = fn(mean(sos_outtime_death / 30.5), dig = 0),
    sd = fn(sd(sos_outtime_death / 30.5, na.rm = T), dig = 0),
    min = fn(min(sos_outtime_death / 30.5), dig = 0),
    max = fn(max(sos_outtime_death / 30.5), dig = 0),
    q1 = fn(quantile(sos_outtime_death / 30.5, probs = 0.25), dig = 0),
    q3 = fn(quantile(sos_outtime_death / 30.5, probs = 0.75), dig = 0)
  ) %>%
  mutate(out = paste0(
    med, " [", min, ", ", q1, ", ",
    q3, ", ", max, "], ", mean, " (", sd, ")"
  )) %>%
  pull(out)

medsurv1 <- rsdata %>%
  filter(shf_indexyear_cat == "2016-2020") %>%
  summarise(
    med = fn(median(sos_outtime_death / 30.55), dig = 0),
    mean = fn(mean(sos_outtime_death / 30.5), dig = 0),
    sd = fn(sd(sos_outtime_death / 30.5, na.rm = T), dig = 0),
    min = fn(min(sos_outtime_death / 30.5), dig = 0),
    max = fn(max(sos_outtime_death / 30.5), dig = 0),
    q1 = fn(quantile(sos_outtime_death / 30.5, probs = 0.25), dig = 0),
    q3 = fn(quantile(sos_outtime_death / 30.5, probs = 0.75), dig = 0)
  ) %>%
  mutate(out = paste0(
    med, " [", min, ", ", q1, ", ",
    q3, ", ", max, "], ", mean, " (", sd, ")"
  )) %>%
  pull(out)

medsurv2 <- rsdata %>%
  filter(shf_indexyear_cat == "2021-2023") %>%
  summarise(
    med = fn(median(sos_outtime_death / 30.55), dig = 0),
    mean = fn(mean(sos_outtime_death / 30.5), dig = 0),
    sd = fn(sd(sos_outtime_death / 30.5, na.rm = T), dig = 0),
    min = fn(min(sos_outtime_death / 30.5), dig = 0),
    max = fn(max(sos_outtime_death / 30.5), dig = 0),
    q1 = fn(quantile(sos_outtime_death / 30.5, probs = 0.25), dig = 0),
    q3 = fn(quantile(sos_outtime_death / 30.5, probs = 0.75), dig = 0)
  ) %>%
  mutate(out = paste0(
    med, " [", min, ", ", q1, ", ",
    q3, ", ", max, "], ", mean, " (", sd, ")"
  )) %>%
  pull(out)
```

The median (q1-q3) age is `r med` and `r percfem`% females.    

The median [minimum, q1, q3, maximum], mean (standard deviation) follow-up time (months) was `r medsurv` for the overall 
population and `r medsurv1` for patients included 2016-2020 and `r medsurv2` for patients included 2021-2023. 

## Baseline characteristics

Baseline characteristics are presented partly by patients’ status, i.e. a HF hospitalization within 3 months previously. 
Categorical variables are presented with n and percentage and tested for differences between groups using the chi-square test. 
Continuous variables are presented with median [minimum, first quartile, third quartile, maximum], mean (standard deviation) and tested for differences between groups using the Kruskal-Wallis test. 

{{< include src/base_tab.qmd >}}

\elandscape

\clearpage

## Primary objective

The primary objective is to describe the frequency of the different co-prescriptions of the drugs among the four pillars for the treatment of HFrEF, before and after the availability of dapagliflozin on the market for an HF indication, according to patients’ status at index date, i.e. patients with recent HF hospitalization (up to 3 months post-discharge) vs stable patients (all the others) and patients’ clinical characteristics at index date.

{{< include src/primary.qmd >}}

\clearpage

\blandscape

## Secondary objectives

### Secondary objective 1

The secondary objective 1 is to assess the association between the number of drugs prescribed among the four pillars (mono, bi, tri or quadruple therapy) and the occurrence of hospitalization for HF and HF hospitalization + CV death according to patient’s status and patients’ clinical characteristics at index date.

{{< include src/sec1.qmd >}}

\clearpage

### Secondary objective 2

The secondary objective 2 is to describe adherence and persistence to the four pillars according to the total number of drugs taken for other CV and non-CV diseases (i.e. overall burden of pharmacotherapy), and considering patients’ status and patient’s clinical characteristics at index date. 

The adherence is calculated as $\ge$ 80% adherence during the first year after index.

{{< include src/sec2.qmd >}}

{{< include src/sec2_disc.qmd >}}

\clearpage

### Secondary objective 3

The secondary objective 3 is to assess the association between the adherence to the four pillars and the risk of HF hospitalization according to patient’s status and patient’s clinical characteristics.

The adherence is calculated as $\ge$ 80% adherence during the first year after index. The follow-up time is started 1 year after baseline when adherence is evaluated. All covariates are from baseline.

{{< include src/sec3.qmd >}}

\clearpage

\blandscape

### Secondary objective 4

The secondary objective 4 is to, in a subset of patients with “de novo” HFrEF (history of HF <6 months) and medication naive for SGLT2i, ACEi/ARB/ARNi, Beta-blocker and MRA (defined as not taking any of these drugs 4-12 months prior to baseline), to describe the initiation of triple or quadruple therapy (sequence), and to assess the associations with the risk of HF hospitalization and HF hospitalization + CV death according to patient’s status and patients’ clinical characteristics at index date.

{{< include src/sec4.qmd >}}

\elandscape

\clearpage

### Secondary objective 5-6

The secondary objective 5 is to describe the number of patients who would be eligible for a SPC containing a SGLT2i, a beta-blocker and an MRA.

The secondary objective 6 is to, among these “eligible” patients, describe the proportion of patients treated by these three classes drugs concomitantly according to patient’s status and patients’ clinical characteristics. Doses will be assessed according to systolic blood pressure, heart rate, renal function, potassium levels and diabetes type I. 

Eligible is defined as: 

- $\ge$ 100 mm Hg or taking any drug, 
- heart rate $\ge$ 50 beats/minute or taking Beta-blocker
- eGFR $\ge$ 30 mL/min/1.73 or taking MRA or SGLT2i
- potassium $\le$ 5.0 mEq/L or taking MRA
- no diabetes type 1 or taking SGLT2i. 

All patients are eligible according to the above definition. 

{{< include src/sec6.qmd >}}

\clearpage

### Secondary objective 7

The secondary objective 7 is to describe the number of patients who would be eligible for a SPC containing the four pillars (SGLT2i, BB, MRA, ACEi/ARNI) 
according to systolic blood pressure, heart rate, renal function, potassium levels and diabetes type I at index. 

Eligible is defined as patients currently on the combination of  drugs. 

{{< include src/sec7.qmd >}}

\clearpage

### Secondary objective 8

The secondary objective 8 is to, among these “eligible” patients, describe the proportion of patients treated by four drug classes concomitantly according to patient’s status and patients’ clinical characteristics.

{{< include src/sec8.qmd >}}

\clearpage

### Secondary objective 9

The secondary objective 9 is to describe the co-prescriptions of other drugs for HFrEF on top of the four pillars according to patients’ status and patient’s clinical characteristics.

{{< include src/sec9.qmd >}}

\clearpage

\blandscape

### Secondary objective 10

The secondary objective 10 is to report the proportion of discontinuation of combinations of the four pillars, and patients´ characteristics associated with discontinuation.

{{< include src/sec10.qmd >}}

\elandscape

\clearpage

### Secondary objective 11

The secondary objective 11 is to report the number of patients who are prescribed dapagliflozin and bisoprolol concomitantly, who would be eligible for a SPC with the combination of these two drugs. 

{{< include src/sec11.qmd >}}

### Secondary objective 12

The secondary objective 12 is to report the number of patients who are prescribed concomitantly dapagliflozin and bisoprolol and remain on these drugs at least six and twelve months, with respective dosage variations. Patients who do not have follow-up until 6 months + 2 months and 1 year + 2 months are excluded from the respective calculations.

{{< include src/sec12.qmd >}}

### Secondary objective 13

The secondary objective 13 is to report the number of patients already on a stable dose of bisoprolol who are not taking dapagliflozin who would be eligible for a SPC with the combination of dapagliflozin and bisoprolol. 

{{< include src/sec13.qmd >}}

\clearpage

### Secondary objective 14

The secondary objective 14 is to report the number  of patients who are simultaneously prescribed dapagliflozin, bisoprolol and an MRA who would be eligible for a SPC with the combination of dapagliflozin, bisoprolol and an MRA.

{{< include src/sec14.qmd >}}

### Add hoc analyses 

{{< include src/adhoc_1.qmd >}}

\clearpage

# Reproducibility

## R code

The R code for all data handling and statistical analyses are found: 

https://github.com/KIHeartFailure/swedehf-hftreat. No data or output is stored in the repository. 

## R session information {#sessioninfo}

```{r sessinfo}
sessionInfo()
```

# References

::: {#refs}
:::
